SSwAB-IT Phase 1 Analysis

### 2024-04-15 Jordana Hoppe/SSwAB-IT/phase_1_analysis.qmd
update for Jordana

```{r setup}
#| include: false
library(redcapAPI)
library(tidyverse)
library(gt)
library(gtsummary)
library(caret)
library(twoxtwo)
library(vcd)
library(eulerr)
```

```{r data cleaning}
#| include: false
# Import from REDCap
unlockREDCap(c(rcon = "SSwAB-IT (New)"),
  keyring = "API_KEYs",
  envir = 1,
  url = "https://redcap.ucdenver.edu/api/"
)
exportBulkRecords(list(db = rcon), envir = 1)
# Exclude
exclude <- c("01-030_W", "01-038_W", "02-0038", "02-0039", "04-001")
# Combine collection results
df <- full_join(db_cos_collection_results, db_sos_collection_results)
# Add dates to demographics
db_demographics <- left_join(db_demographics,
  df %>% select(record_id, cos_result_date),
  by = join_by(record_id)
)
db_demographics$age_at_cos <- difftime(db_demographics$cos_result_date,
  db_demographics$dob,
  units = "days"
) / 365.25
# Format
df <- df %>%
  select(
    record_id, redcap_data_access_group,
    staphylococcus_aureus_cos:other_result_cos,
    staphylococcus_aureus_sos:other_result_sos
  ) %>%
  filter(!record_id %in% exclude)
# Remove empty demographics rows
db_demographics <- db_demographics %>%
  filter(
    redcap_event_name == "Clinic Visit (Arm 1: Eligible)",
    !record_id %in% exclude
  )
# Excluded participants
db_medications <- db_medications %>% filter(!record_id %in% exclude)
# Combine race and ethnicity columns
races <- c(
  "American Indian or Alaskan Native", "Asian", "Hawaiian or Pacific Islander",
  "Black or African American", "White", "Unknown", "Other"
)
ethnicities <- c("Hispanic", "Non-Hispanic", "Unknown/Not Reported")
db_demographics$Race <- apply(db_demographics, 1, function(r) {
  w <- which(r[grep("race", names(r))] == "Checked")
  if (length(w) == 1) {
    return(races[w])
  } else if (length(w) > 1) {
    return("Multiple Races")
  } else if (length(w) == 0) {
    return(NA)
  }
})
db_demographics$Ethnicity <- apply(db_demographics, 1, function(r) {
  w <- which(r[grep("ethnicity", names(r))] == "Checked")
  if (length(w) == 1) {
    return(ethnicities[w])
  } else if (length(w) > 1) {
    return(paste(ethnicities[w], sep = "/"))
  } else if (length(w) == 0) {
    return(NA)
  }
})
# # Drop levels
# level_drop <- c("sex", "Race", "Ethnicity")
# db_demographics[, level_drop] <- lapply(
#   db_demographics[, level_drop],
#   function(c) {
#     droplevels(factor(c))
#   }
# )
# Add groups to other dataframes
db_medications <- left_join(db_medications,
  db_demographics %>% select(record_id, group),
  by = join_by(record_id)
)
# Bug list
bugs <- c(
  "staphylococcus_aureus", "haemophilus_influenza", "pseudomonas_aeruginosa",
  "pseudomonas_non_aero", "s_maltophilia", "burkholderia", "aspergillus", "ntm",
  "other"
)
bug_names <- c(
  "S. aureus", "H. influenza", "P. aeruginosa", "Pseudomonas non-aeruginosa",
  "S. maltophilia", "Burkholderia", "Aspergillus", "NTM", "Other"
)
# Convert yes/no to positive/negative
df[, paste0(bugs, "_cos")] <- lapply(df[, paste0(bugs, "_cos")], factor,
  levels = c("No", "Yes"), labels = c("Negative", "Positive")
)
df[, paste0(bugs, "_sos")] <- lapply(df[, paste0(bugs, "_sos")], factor,
  levels = c("No", "Yes"), labels = c("Negative", "Positive")
)
# Create overall columns (any bug detected in sample)
df$overall_cos <- apply(df, 1, function(r) {
  any(r[paste0(bugs, "_cos")] == "Positive")
})
df$overall_cos <- factor(df$overall_cos,
  levels = c(F, T), labels = c("Negative", "Positive")
)
df$overall_sos <- apply(df, 1, function(r) {
  any(r[paste0(bugs, "_sos")] == "Positive")
})
df$overall_sos <- factor(df$overall_sos,
  levels = c(F, T), labels = c("Negative", "Positive")
)
# Add overall to list
bugs <- c("overall", bugs)
bug_names <- c("Across all bugs", bug_names)
bug_list <- as.list(bug_names)
names(bug_list) <- bugs
# Agreement columns
df[, bugs] <- lapply(bugs, function(b) {
  cos <- df[, paste0(b, "_cos")]
  sos <- df[, paste0(b, "_sos")]
  f <- factor(paste(cos, sos, sep = "_"),
    exclude = c(
      "NA_Negative", "Negative_NA", "NA_Positive", "Positive_NA", "NA_NA"
    )
  )
  f <- factor(f,
    levels = c(
      "Negative_Negative", "Negative_Positive", "Positive_Negative",
      "Positive_Positive"
    ),
    labels = c(
      "COS Negative / SOS Negative", "COS Negative / SOS Positive",
      "COS Positive / SOS Negative", "COS Positive / SOS Positive"
    )
  )
  return(f)
})
```

@@

```{r}
db_demographics %>%
  select(group, age_at_cos, sex, Race, Ethnicity) %>%
  tbl_summary(
    by = group,
    statistic = all_continuous()~c("{median} ({min}, {max})"),
    label = list("sex" = "Sex", "age_at_cos" = "Age at COS")
  ) %>%
  add_overall() %>%
  add_p(test = list(c("Race", "Ethnicity") ~ "fisher.test")) %>%
  separate_p_footnotes() %>%
  as_gt()
```

@@

Bangdiwala's agreement plot @bangdiwalaAgreementChart2013 is a way of visualizing 2 x 2 contingency tables, and additional details are in the paper referenced at the end of this document.

```{r}
agreement <- function(bug, data = df) {
  cos <- paste0(bug, "_cos")
  sos <- paste0(bug, "_sos")
  d <- data[, c(cos, sos)]
  d <- d[complete.cases(d), ]
  # Agreement metrics
  c <- confusionMatrix((d[, sos]),
    reference = d[, cos], positive = "Positive", mode = "everything",
    dnn = c("SOS", "COS")
  )
  # Return
  return(list(two_by_two = c$table, accuracy = c$overall))
}
```

```{r}
l <- lapply(bugs, function(b) {
  a <- agreement(b)
  # Remove those where both tests were positive
  n <- a$two_by_two
  n[2, 2] <- 0
  # Proportion tests for non-agreement in participants with one negative test
  cos_p <- prop.test(n[1, 2], sum(n))
  sos_p <- prop.test(n[2, 1], sum(n))
  nn <- prop.test(n[1, 1], sum(n))
  return(list(
    Accuracy = a$accuracy["Accuracy"],
    `Accuracy LL` = a$accuracy["AccuracyLower"],
    `Accuracy UL` = a$accuracy["AccuracyUpper"],
    Kappa = a$accuracy["Kappa"],
    `Mcnemar P Value` = a$accuracy["McnemarPValue"],
    `N With Negative Test on Either` = sum(n),
    `N Negative on Both` = n[1, 1],
    `Prop. Both Negative` = nn$estimate,
    `Prop. Both Negative LL ` = nn$conf.int[1],
    `Prop. Both Negative UL ` = nn$conf.int[2],
    `N Postive on COS and Negative on SOS` = n[1, 2],
    `Prop. Postive on COS` = cos_p$estimate,
    `Prop. Postive on COS LL ` = cos_p$conf.int[1],
    `Prop. Postive on COS UL ` = cos_p$conf.int[2],
    `N Postive on SOS and Negative on COS` = n[2, 1],
    `Prop. Postive on SOS` = sos_p$estimate,
    `Prop. Postive on SOS LL ` = sos_p$conf.int[1],
    `Prop. Postive on SOS UL ` = sos_p$conf.int[2]
  ))
})
l <- data.frame(do.call(rbind, l), check.names = F)
l <- data.frame(lapply(l, as.numeric), check.names = F)
rownames(l) <- bug_names
gt(l, rownames_to_stub = T) %>% fmt_number(n_sigfig = 3)
```

```{r results='asis'}
invisible(lapply(names(bug_list), function(b) {
  cat("\n")
  cat("##", as.character(bug_list[b]))
  cat("\n")
  cos <- paste0(b, "_cos")
  sos <- paste0(b, "_sos")
  d <- df[, c(cos, sos)]
  d <- d[complete.cases(d), ]
  colnames(d) <- c("COS", "SOS")
  cat("\n")
  c <- confusionMatrix(d[, "SOS"],
    reference = d[, "COS"], positive = "Positive", mode = "everything"
  )
  metrics <- data.frame(c$overall)
  cat("### 2x2 table")
  cat("\n")
  two_by_two <- data.frame(c$table)
  print(gt(two_by_two))
  cat("\n")
  cat("### Accuracy")
  cat("\n")
  print(gt(metrics, rownames_to_stub = T) %>%
    fmt_number(n_sigfig = 3) %>%
    tab_options(column_labels.hidden = TRUE))
  cat("\n")
  cat("### Agreement plot")
  cat("\n")
  print(agreementplot(table(d)))
  cat("\n")
  cat("### Positive/Positive Excluded")
  cat("\n")
  t <- d[rowSums(d == "Positive") < 2, ]
  t$`Samples With a Negative Test` <- paste(
    "COS", t[, "COS"], "/", "SOS", t[, "SOS"]
  )
  t %>%
    select(`Samples With a Negative Test`) %>%
    tbl_summary() %>%
    add_ci() %>%
    as_gt() %>%
    print()
  cat("\n")
  cat("### Euler diagram of negative results")
  cat("\n")
  print(plot(euler(d == "Negative"), quantities = T))
  cat("\n")
}))
```

>>>>

```{r setup}
#| include: false
library(redcapAPI)
library(tidyverse)
library(gt)
library(gtsummary)
library(caret)
library(eulerr)
```

```{r data cleaning}
#| include: false
# Import from REDCap
unlockREDCap(c(rcon = "SSwAB-IT (New)"),
  keyring = "API_KEYs",
  envir = 1,
  url = "https://redcap.ucdenver.edu/api/"
)
exportBulkRecords(list(db = rcon), envir = 1)
# Exclude
exclude <- c("01-030_W", "01-038_W", "02-0038", "02-0039", "04-001")
# Combine collection results
df <- full_join(db_cos_collection_results, db_sos_collection_results)
# Add dates to demographics
db_demographics <- left_join(db_demographics,
  df %>% select(record_id, cos_result_date),
  by = join_by(record_id)
)
db_demographics$age_at_cos <- as.numeric(difftime(db_demographics$cos_result_date,
  db_demographics$dob,
  units = "days"
)) / 365.25
db_demographics$age_group <- cut(db_demographics$age_at_cos, c(-Inf, 18, Inf),
  right = F
)
# Format
df <- df %>%
  select(
    record_id, redcap_data_access_group,
    staphylococcus_aureus_cos:other_result_cos,
    staphylococcus_aureus_sos:other_result_sos
  ) %>%
  filter(!record_id %in% exclude)
# Remove empty demographics rows
db_demographics <- db_demographics %>%
  filter(
    redcap_event_name == "Clinic Visit (Arm 1: Eligible)",
    !record_id %in% exclude
  )
# Excluded participants
db_medications <- db_medications %>% filter(!record_id %in% exclude)
# Combine race and ethnicity columns
races <- c(
  "American Indian or Alaskan Native", "Asian", "Hawaiian or Pacific Islander",
  "Black or African American", "White", "Unknown", "Other"
)
ethnicities <- c("Hispanic", "Non-Hispanic", "Unknown/Not Reported")
db_demographics$Race <- apply(db_demographics, 1, function(r) {
  w <- which(r[grep("race", names(r))] == "Checked")
  if (length(w) == 1) {
    return(races[w])
  } else if (length(w) > 1) {
    return("Multiple Races")
  } else if (length(w) == 0) {
    return(NA)
  }
})
db_demographics$Ethnicity <- apply(db_demographics, 1, function(r) {
  w <- which(r[grep("ethnicity", names(r))] == "Checked")
  if (length(w) == 1) {
    return(ethnicities[w])
  } else if (length(w) > 1) {
    return(paste(ethnicities[w], sep = "/"))
  } else if (length(w) == 0) {
    return(NA)
  }
})
# Add groups to other dataframes
db_medications <- left_join(db_medications,
  db_demographics %>% select(record_id, group),
  by = join_by(record_id)
)
# Bug list
bugs <- c(
  "staphylococcus_aureus", "haemophilus_influenza", "pseudomonas_aeruginosa",
  "pseudomonas_non_aero", "s_maltophilia", "burkholderia", "aspergillus", "ntm",
  "other"
)
bug_names <- c(
  "S. aureus", "H. influenza", "P. aeruginosa", "Pseudomonas non-aeruginosa",
  "S. maltophilia", "Burkholderia", "Aspergillus", "NTM", "Other"
)
# Convert yes/no to positive/negative
df[, paste0(bugs, "_cos")] <- lapply(df[, paste0(bugs, "_cos")], factor,
  levels = c("No", "Yes"), labels = c("Negative", "Positive")
)
df[, paste0(bugs, "_sos")] <- lapply(df[, paste0(bugs, "_sos")], factor,
  levels = c("No", "Yes"), labels = c("Negative", "Positive")
)
# Create overall columns (any bug detected in sample)
df$overall_cos <- apply(df, 1, function(r) {
  any(r[paste0(bugs, "_cos")] == "Positive")
})
df$overall_cos <- factor(df$overall_cos,
  levels = c(F, T), labels = c("Negative", "Positive")
)
df$overall_sos <- apply(df, 1, function(r) {
  any(r[paste0(bugs, "_sos")] == "Positive")
})
df$overall_sos <- factor(df$overall_sos,
  levels = c(F, T), labels = c("Negative", "Positive")
)
# Add overall to list
bugs <- c("overall", bugs)
bug_names <- c("Across all bugs", bug_names)
bug_list <- as.list(bug_names)
names(bug_list) <- bugs
# Add group to dataframe
df <- left_join(df, db_demographics %>% select(record_id, group),
  by = join_by(record_id)
)
```

@@

```{r}
db_demographics %>%
  select(group, age_at_cos, age_group, sex, Race, Ethnicity) %>%
  tbl_summary(
    by = group,
    statistic = all_continuous() ~ c("{median} ({min}, {max})"),
    label = list(
      "sex" = "Sex", "age_at_cos" = "Age at COS", "age_group" = "Age Group"
    )
  ) %>%
  add_overall() %>%
  add_p(test = list(c("Race", "Ethnicity") ~ "fisher.test")) %>%
  separate_p_footnotes() %>%
  as_gt()
```

@@

```{r}
agreement <- function(bug, data = df) {
  cos <- paste0(bug, "_cos")
  sos <- paste0(bug, "_sos")
  d <- data[, c(cos, sos)]
  colnames(d) <- c("COS", "SOS")
  d <- d[complete.cases(d), ]
  # Agreement metrics
  c <- confusionMatrix(d$SOS,
    reference = d$COS,
    positive = "Positive", mode = "everything",
    dnn = c("SOS", "COS")
  )
  # Print
  cat("\n")
  cat("####", "2x2 Table")
  cat("\n")
  d %>%
    tbl_cross(percent = "cell") %>%
    print()
  cat("\n")
  cat("####", "2x2 Table (With CIs)")
  cat("\n")
  d %>%
    tbl_cross(percent = "cell") %>%
    add_ci() %>%
    print()
  cat("\n")
  cat("Overall accuracy (both negative/negative and positive/positive) was ",
    scales::percent(c$overall["Accuracy"], accuracy = 0.1), " (95% CI: ",
    scales::percent(c$overall["AccuracyLower"], accuracy = 0.1), " - ",
    scales::percent(c$overall["AccuracyUpper"], accuracy = 0.1), ").",
    sep = ""
  )
  cat("\n")
  cat("\n")
  cat("Kappa was ", round(c$overall["Kappa"], 3), ".", sep = "")
  cat("\n")
  cat("\n")
  cat("#### Euler diagram of negative results")
  cat("\n")
  print(plot(euler(d == "Negative"), quantities = T))
  cat("\n")
  cat("\n")
  cat("#### Euler diagram of positive results")
  cat("\n")
  print(plot(euler(d == "Positive"), quantities = T))
  cat("\n")
}
agreement_test <- function(bug, group = "group", data = df) {
  # Prepare data
  cos <- paste0(bug, "_cos")
  sos <- paste0(bug, "_sos")
  d <- data[, c(cos, sos, group)]
  d <- d[complete.cases(d), ]
  colnames(d) <- c("COS", "SOS", "Group")
  # Agreement and detailed agreement
  d$agree <- d$COS == d$SOS
  d$detailed_agree <- paste0("COS ", d$COS, "/", "SOS ", d$SOS)
  # Print
  cat("\n")
  d %>%
    select(agree, detailed_agree, Group) %>%
    tbl_summary(by = Group, label = list(
      agree ~ "Agree (Both Positive/Positive and Negative/Negative)",
      detailed_agree ~ "Detailed Agreement"
    )) %>%
    add_overall() %>%
    add_p(test.args = all_tests("fisher.test") ~ list(workspace = 2e6)) %>%
    separate_p_footnotes() %>%
    print()
  cat("\n")
}
```

```{r results='asis'}
invisible(lapply(names(bug_list), function(b) {
  cat("\n")
  cat("##", as.character(bug_list[b]))
  cat("\n")
  cat("###", "Overall agreement")
  cat("\n")
  agreement(b)
  cat("\n")
  cat("###", "Agreement by group")
  cat("\n")
  agreement_test(b)
  cat("\n")
}))

# invisible(lapply(names(bug_list), function(b) {
#   cat("\n")
#   cat("##", as.character(bug_list[b]))
#   cat("\n")
#   cos <- paste0(b, "_cos")
#   sos <- paste0(b, "_sos")
#   d <- df[, c(cos, sos)]
#   d <- d[complete.cases(d), ]
#   colnames(d) <- c("COS", "SOS")
#   cat("\n")
#   c <- confusionMatrix(d[, "SOS"],
#     reference = d[, "COS"], positive = "Positive", mode = "everything"
#   )
#   metrics <- data.frame(c$overall)
#   cat("### 2x2 table")
#   cat("\n")
#   two_by_two <- data.frame(c$table)
#   print(gt(two_by_two))
#   cat("\n")
#   cat("### Accuracy")
#   cat("\n")
#   print(gt(metrics, rownames_to_stub = T) %>%
#     fmt_number(n_sigfig = 3) %>%
#     tab_options(column_labels.hidden = TRUE))
#   cat("\n")
#   cat("### Agreement plot")
#   cat("\n")
#   print(agreementplot(table(d)))
#   cat("\n")
#   cat("### Positive/Positive Excluded")
#   cat("\n")
#   t <- d[rowSums(d == "Positive") < 2, ]
#   t$`Samples With a Negative Test` <- paste(
#     "COS", t[, "COS"], "/", "SOS", t[, "SOS"]
#   )
#   t %>%
#     select(`Samples With a Negative Test`) %>%
#     tbl_summary() %>%
#     add_ci() %>%
#     as_gt() %>%
#     print()
#   cat("\n")
#   cat("### Euler diagram of negative results")
#   cat("\n")
#   print(plot(euler(d == "Negative"), quantities = T))
#   cat("\n")
# }))
```
